Adaptive global optimization using graphics accelerators\thanks{This study was supported by the Russian Science Foundation, project No.\,16-11-10150.}

### 2020-04-13 2020/RSD_GPU/RSD2020.tex
Fix some mistakes

\documentclass{svproc}
%
% RECOMMENDED 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%
%
\usepackage{graphicx}
\usepackage{marvosym}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[russian]{babel}

@@

\begin{document}
\mainmatter              % start of a contribution
%
\title{Adaptive global optimization using graphics accelerators\thanks{This study was supported by the Russian Science Foundation, project No.\,16-11-10150.}
}
%
\titlerunning{Adaptive global optimization}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{Konstantin Barkalov\letter\orcidID{0000-0001-5273-2471} \and \\  Ilya Lebedev\orcidID{0000-0002-8736-0652}}
%
\authorrunning{Konstantin Barkalov \and Ilya Lebedev} % abbreviated author list (for running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
\tocauthor{Konstantin Barkalov and Ilya Lebedev}
%
\institute{Lobachevsky State University of Nizhni Novgorod, Russia  \\
	\email{konstantin.barkalov@itmm.unn.ru},
	\email{ilya.lebedev@itmm.unn.ru}
}

%\maketitle              % typeset the title of the contribution

@@

A promising direction in the field of parallel global optimization (which, indeed, is true in many areas related to the software implementation of time-consuming algorithms) is the use of graphics processing units (GPUs). In the past decade, graphics accelerators have rapidly increased performance to meet the ever-growing demands of graphics application developers. Additionally, in the past few years some principles for developing graphics hardware have changed, and as a result it has become more programmable. Today, a graphics accelerator is a flexibly programmable, massive parallel processor with high performance, which is in demand for solving a range of computationally time-consuming problems \cite{Ref1}.

However, the potential for graphics accelerators to solve global optimization problems has not yet been fully realized. Using the GPUs, they basically parallelize nature-inspired optimization algorithms, which are somehow based on the idea of random search (see, for example, \cite{Ref2,Ref3,Ref4}). By virtue of their stochastic nature, algorithms of this type guarantee convergence to the global minimum only in the sense of probability, which differentiates them unfavorably from deterministic methods.

With regard to many deterministic algorithms of Lipschitzian global optimization with guaranteed convergence, parallel variants have been proposed \cite{Ref5,Ref6,Ref7}. However, these versions of algorithms are parallelized on the CPU using shared and/or distributed memory; presently, no GPU implementations have been made. For example, \cite{Ref7} describes parallelization of an algorithm based on the ideas of the branch and boundary method using MPI and OpenMP.

Within the framework of this research, we consider the problems of capturing the optimum, which are characterized by a lengthy period for calculating the values of objective function in comparison with the time needed for processing them. For example, objective function can be specified using systems of linear algebraic equations, systems of ordinary differential equations, etc. Currently, graphics accelerators can be used to solve problems of this type. Moreover, an accelerator can solve several such problems at once \cite{Ref8}; i.e., using the GPU, one can calculate multiple function values simultaneously.

@@

\section{Multidimensional parallel global search algorithm}
Consider the problem of finding the global minimum of the \(N\)-dimensional function \(\phi(y)\) in the hyperinterval  \(D=\{y\in R^N:a_i\leqslant x_i\leqslant{b_i}, 1\leqslant{i}\leqslant{N}\}\)  We will assume that the function satisfies the Lipschitz condition with an a priori unknown constant \(L\).
\begin{equation}
\label{task}
\phi(y^*)=\min\{\phi(y):y\in D\}
\end{equation}
\begin{equation}
\label{lip}
|\phi(y_1)-\phi(y_2)|\leqslant L\Vert y_1-y_2\Vert,y_1,y_2\in D,0<L<\infty
\end{equation}

@@

The questions of numerical construction of approximations to Piano curve (evolvents) and the corresponding theory are discussed in detail in \cite{Ref9}, \cite{Ref10}. Using evolvents \(y(x)\) reduces the multidimensional problem (\ref{task}) to a one-dimensional problem
\begin{displaymath}
\label{oneDimTask}
\phi(y^*)=\phi(y(x^*))=\min\{\phi(y(x)):x\in [0,1]\}
\end{displaymath}

An important property is that the relative differences of the function remain limited: if the function \(\phi(y)\) in the region \(D
\) satisfies the Lipschitz condition, then the function \(\phi(y(x))\) in the interval \([0,1]\) will satisfy the uniform Holder condition
\begin{displaymath}
\label{holder}
|\phi(y(x_1))-\phi(y(x_2))|\leqslant H{|x_1-x_2|}^{\frac{1}{N}},x_1,x_2\in[0,1]
\end{displaymath}
where the Holder constant \(H\) is related to the Lipschitz constant \(L\) by the ratio
\begin{displaymath}
H=4Ld\sqrt{N},d=\max\{b_i-a_i:1\leqslant i\leqslant N\}
\end{displaymath}

Therefore, without limiting generality, we can consider minimizing the one-dimensional function \(f(x)=\phi(y(x)), x\in[0,1]\), which satisfies the Holder condition.

@@

instead of using a single Peano curve \(y(x)\) (see \cite{Ref19}, \cite{Ref20}). For example, each Peano curve \(y^i(x)\) from \(Y_L(x)\) can be obtained as a result of some shift along the main diagonal of the hyperinterval \(D\), the original curve \(y(x)\). Another way is to rotate the sweep around the coordinate. The set of evolvents that have been constructed allows us to obtain for any close images \(y', y''\), close prototypes \(x', x''\) for some mapping \(y^i(x)\).

@@

Each problem from this set can be solved independently, and any calculated value \(z= \varphi(y')\), \(y'=y^i(x')\) of the function \(\varphi(y)\) in the \(i\)-th problem can be interpreted as calculating the value \(z= \varphi(y')\), \(y'=y^s(x'')\) for any other \(s\)-th problem without repeated labor-intensive calculations of the function \(\varphi(y)\). Such informational unity makes it possible to solve the entire set of problems in a parallel fashion. This approach was discussed in detail in \cite{Ref21}.

@@

The recursive optimization scheme is based on the well-known relation \cite{Ref22}

\begin{equation}
\label{nestedScheme}
\min{\phi(y):y\in D}=\min_{a_1\leqslant y_1\leqslant b_1}\min_{a_2\leqslant y_2\leqslant b_2}\dots\min_{a_1\leqslant y_N\leqslant b_N}\phi(y)
\end{equation}

which allows one to replace the solution of the multidimensional problem (\ref{task}) with the solution of a family of one-dimensional subproblems recursively related to each other.
Let's introduce a set of functions

\begin{equation}
\label{phiN}
\phi_N(y_1,\dots,y_N)=\phi(y_1,\dots,y_N)
\end{equation}
\begin{equation}
\label{phiI_}
\phi_i(y_1,\dots,y_i)=\min_{a_{i+1}\leqslant y_{i+1} \leqslant b_{i+1}}\phi_{i+1}(y_1,\dots,y_i,y_{i+1}),1\leqslant i\leqslant N-1
\end{equation}

@@

\begin{equation}
\label{phiFirst}
\phi_1(y_1^*)=\min\{\phi_1(y_1),y_1\in [a,b]\}
\end{equation}

However, each calculation of the value of the one-dimensional function  \(\phi_1(y_1)\) at some fixed point corresponds to the solution of a one-dimensional minimization problem

\begin{displaymath}
\label{phi2}
\phi_2(y_1,y_2^*)=\min\{\phi(y_1,y_2):y_2\in [a_2,b_2]\}
\end{displaymath}

@@

For the recursive scheme described above, a generalization (block recursive scheme) is proposed that combines the use of sweeps and a recursive scheme in order to efficiently parallelize computations.

Consider the vector y as a vector of block variables
\begin{displaymath}
y=(y_1,\dots,y_N)=(u_1,u_2,\dots,u_M)
\end{displaymath}

Where the \(i\)-th block variable \(u_i\) is a dimensional vector   of sequentially taken components of the vector \(y\), i.e. \(u_1=(y_1,y_2,\dots,y_{N_1}),u_2=(y_{N_1+1},y_{N_1+2},\dots,y_{N_1+N_2}),\dots,u_M=(y_{N-N_M+1},y_{N-N_M+2},\dots,y_N)\), while \(N_1+N_2+\dots+N_M=N\).

@@

\begin{equation}
\label{blockNested}
\min\phi(y)_{y\in D}=\min_{u_1\in D_1}\min_{u_2\in D_2}\dots\min_{u_M\in D_M}\phi(y)
\end{equation}

@@

\begin{equation}
\label{subTasks}
\phi_i(u_1,\dots,u_i)=\min_{u_{i+1}\in D_{i+1}}\phi_{i+1}(u_1,\dots,u_i,u_{i+1}),1\leqslant i\leqslant M-1
\end{equation}

@@

The obvious method (elaborated in detail in \cite{Ref23},\cite{Ref24},\cite{Ref25} for the nested optimization scheme and in \cite{Ref26},\cite{Ref27} for the block nested optimization scheme) is based on solving subproblems in accordance with the recursive order of their generation. However, there is a significant loss of information about the target function.

@@

In the case of one-dimensional subproblems, this approach was theoretically substantiated and tested in \cite{Ref28}, \cite{Ref29}, \cite{Ref30}, and in the article [] a generalization of the adaptive scheme for multidimensional subproblems was proposed.

@@

Let the nested subproblems in the form (\ref{subTasks}) be solved using the global search algorithm described in subsection 1. Then each subproblem (\ref{subTasks}) can be assigned a numerical value called the characteristic of this problem. As such, we can take the maximum characteristic of the intervals formed in this problem – \(R(t)\). In accordance with the rule for calculating characteristics, the higher the value of the characteristic, the more promising the subproblem is in the continued search for the global minimum of the original problem (\ref{task}). Therefore, at each iteration, subproblems with the maximum characteristic are selected for conducting the next test. The test either leads to the calculation of the value of the objective function \(\phi_1(y)\) (if the selected subproblem belonged to the level \(j=M\)), or generates new subproblems according to (\ref{subTasks}) for \(j \geqslant  M-1\). In the latter case, the newly generated problems are added to the current set of problems, their characteristics are calculated, and the process is repeated. The optimization process is completed when the root problem satisfies the condition for stopping the algorithm that solves this problem. Some results pointing in this direction are presented in \cite{Ref31}, \cite{Ref32}.

@@

\ref{fig:Parallel_Computing} shows the general scheme of organizing calculations using several cluster nodes and several GPUs. In accordance with this scheme, nested subproblems  ${{\phi }_{i}}({{u}_{1}},...,{{u}_{i}})=\underset{{{u}_{i+1}}\in {{D}_{i+1}}}{\mathop{\min }}\,{{\phi }_{i+1}}({{u}_{1}},...,{{u}_{i}},{{u}_{i+1}})$  with \(i=1,\ldots ,M-2\) are solved using only the CPU. The values of the optimized function are not calculated directly in these subproblems: the calculation the function value  ${{\phi }_{i}}({{u}_{1}},...,{{u}_{i}})$ is a solution to the minimization problem at the next level.

The subproblem of the last \((M - 1)\) level \({{\phi }_{M-1}}({{u}_{1}},...,{{u}_{M-1}})=\underset{{{u}_{M}}\in {{D}_{M}}}{\mathop{\min }}\,{{\phi }_{M}}({{u}_{1}},...,{{u}_{M}})\) differs from all the previous subproblems – it calculates the values of the optimized function, since ${{\phi }_{M}}({{u}_{1}},...,{{u}_{M}})=\phi ({{y}_{1}},...,{{y}_{N}})$. This subproblem transfers between the CPU and the GPU.

@@

In the study \cite{Ref15}, \cite{Ref16}, a GKLS generator is described that allows one to generate multiextremal optimization problems with previously known properties: the number of local minima, the size of their regions of attraction, the global minimum point, the value of the function in it, etc.

Below are the results of a numerical comparison of three sequential algorithms -- DIRECT \cite{Ref15}, DIRECTl \cite{Ref17} and Global Search Algorithm (GSA); (results of the first two algorithms are given in the article \cite{Ref18}). A numerical comparison was carried out on the Simple and Hard function classes of dimension 4 and 5 from \cite{Ref18}. The Global minimum $y*$ was considered found if the algorithm generated a test point $y^k$ in the $\delta$-neighborhood of the global minimum, i.e. $\left\| {{y}^{k}}-{{y}^{*}} \right\|\le \delta $. Also, the size of the neighborhood was chosen (in accordance with \cite{Ref18}) as $\delta =\left\| b-a \right\|\sqrt[N]{\Delta }$, here $N$ is the dimension of the problem to be solved, a и b are the boundaries of the search domain $D$, the parameter  $\Delta $=10-6 for N=4 and $\Delta =10-7$ for $N=5$. When using the AGP method for the Simple class, the parameter $r=4.5$ was selected, for the Hard class - $r=5.6$; the parameter for constructing the Peano curve was fixed, $m=10$. The maximum number of iterations allowed was $Kmax = 1 000 000$.

\begin{table}
    \centering
    \begin{tabular}{|c|p{3cm}|c|c|c|}
    \hline
    \(N\) & Problem class & DIRECT & DIRECT\(l\) & GSA \\ \hline
    4 & Simple & >47282 (4) & 18983 & 11953\\ \cline{2-5}
      & Hard & >95708 (7) & 68754 & 25263\\ \hline
    5 & Simple & >16057 (1) & 16758 & 15920\\ \cline{2-5}
      & Hard & >217215 (16) & >269064 (4) & >148342 (4)\\ \hline
    \end{tabular}
    \caption{Average number of iterations}
    \label{table:average_iters}
\end{table}

   Table  \ref{table:average_iters} shows the average number of iterations, $kav$, that the method performed when solving a series of problems from these classes. The symbol ">" reflects a situation where not all problems of the class were solved by any method whatsoever. This means that the algorithm was halted because the maximum allowed number of $Kmax$ iterations was reached. In this case, the value $Kmax = 1 000 000$ was used to calculate the average value of the number of iterations, $kav$, which corresponds to the lower estimate of this average value. The number of unsolved problems is indicated in parentheses.

  As can be seen from Table \ref{table:average_iters}, the sequential GSA surpasses the DIRECT and DIRECTl methods in all classes of problems in terms of the average number of iterations. At the same time, in the 5-Hard class, none of the methods solved all the problems: DIRECT failed to solve 16 problems, DIRECTl and GSA -- 4 problems each.

@@

Следующий эксперимент будет проведен на 6-ти мерных задачах GKLS. В таблице \ref{table:GKLS6_average_MMAS_GPU} приведено ускорение по времени и по итерациям полученное при решении серии задач на GPU с использованием алгоритма объединяющего множественные развертки и адаптивную схему по сравнению с адаптивной схемой на CPU использующей 4 потока. Эксперимент проводился на трех узлах кластера и спользовалось до 6144 нитей.

@@

%
% ---- Bibliography ----
%
\begin{thebibliography}{6}

\bibitem {Ref1}	Hwu, W. GPU Computing Gems: Emerald Edition — San Francisco: Morgan Kaufmann, 2011. — 886 p.
\bibitem {Ref2}	A.M. Ferreiro, J.A. Garcia, J.G. Lopez-Salas, C. Vazquez. An efficient implementation of parallel simulated annealing algorithm in GPUs // Journal of global optimization. — 2013. — V. 57, No 3. — P. 863–890.
\bibitem {Ref3}	J.M. Garcia-Martinez, E.M. Garzon, P.M. Ortigosa. A GPU implementation of a hybrid evolutionary algorithm: GPuEGO // Journal of super-computing. — 2014. — V. 70, No 2. — P. 684–695.
\bibitem {Ref4}	Langdon, W.B. Graphics processing units and genetic programming: an overview // Soft Computing. — 2011. — V. 15, No 8. — P. 1657–1669.
\bibitem {Ref5}	Ю.Г. Евтушенко, В.У. Малкова, А.А. Станевичюс. Параллельный поиск глобального экстремума функций многих переменных // Ж. вычисл. матем. и матем. физ. — 2009. — Т. 49, No2. — С. 255–269.
\bibitem {Ref6}	J. He, A. Verstak, L.T. Watson, M. Sosonkina. Design and implementation of a massively parallel version of DIRECT // Computational optimization and applications. — 2008. — V. 40, No 2. — P. 217–245.
\bibitem {Ref7}	R. Paulavicius, J. Žilinskas, A. Grothey. Parallel branch and bound for global optimization with combination of Lipschitz bounds // Optimization methods and software. — 2011. — V. 26, No 3. — P. 487–498.
\bibitem {Ref8}	V. Kindratenko (ed.). Numerical computations with GPUs — New York: Springer, 2014. — 405 p.
\bibitem {Ref9}	Strongin, R.G. Global Optimization with Non-convex Constraints. Sequential and Parallel Algorithms / R.G. Strongin, Ya.D. Sergeyev – Kluwer Academic Publishers, 2000. – 704 p.
\bibitem {Ref10}	Sergeyev, Ya.D. Introduction to global optimization exploiting space-filling curves / Ya.D. Sergeyev, R.G. Strongin, D. Lera – Springer, 2013. – 125 p.
\bibitem {Ref11}	Стронгин, Р.Г. Параллельные вычисления в задачах глобальной оптимизации / Р.Г. Стронгин, В.П. Гергель, В.А. Гришагин, К.А. Баркалов – М.: Издательство Московского университета, 2013. – 280 с.
\bibitem {Ref12}	Barkalov, K.A. A global optimization technique with an adaptive order of checking for constraints / K.A. Barkalov, R.G. Strongin // Computational Mathematics and Mathematical Physics. – 2002. – Vol. 42, No. 9. – P. 1289-1300.
\bibitem {Ref13}	Gergel, V.P. A method of using derivatives in the minimization of multiextremum functions / V.P. Gergel // Computational Mathematics and Mathematical Physics. – 1996. – Vol. 36, No. 6. – P. 729 – 742.

>>>>

\documentclass{svproc}
%
% RECOMMENDED 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%
%
\usepackage{graphicx}
\usepackage{marvosym}
\usepackage{amsmath}
\usepackage{amssymb}

%\usepackage[russian]{babel}

@@

\begin{document}
\mainmatter              % start of a contribution
%
\title{Adaptive global optimization using graphics accelerators\thanks{This study was supported by the Russian Science Foundation, project No.\,16-11-10150.}
}
%
\titlerunning{Adaptive global optimization}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{Konstantin Barkalov\letter%\orcidID{0000-0001-5273-2471} 
\and Ilya Lebedev%\orcidID{0000-0002-8736-0652}
}

%
\authorrunning{Konstantin Barkalov \and Ilya Lebedev} % abbreviated author list (for running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
\tocauthor{Konstantin Barkalov and Ilya Lebedev}
%
\institute{Lobachevsky State University of Nizhni Novgorod, Russia  \\
	\email{konstantin.barkalov@itmm.unn.ru},
	\email{ilya.lebedev@itmm.unn.ru}
}

\maketitle              % typeset the title of the contribution

@@

A promising direction in the field of parallel global optimization (which, indeed, is true in many areas related to the software implementation of time-consuming algorithms) is the use of graphics processing units (GPUs). In the past decade, graphics accelerators have rapidly increased performance to meet the ever-growing demands of graphics application developers. Additionally, in the past few years some principles for developing graphics hardware have changed, and as a result it has become more programmable. Today, a graphics accelerator is a flexibly programmable, massive parallel processor with high performance, which is in demand for solving a range of computationally time-consuming problems \cite{Hwu2011}.

However, the potential for graphics accelerators to solve global optimization problems has not yet been fully realized. Using the GPUs, they basically parallelize nature-inspired optimization algorithms, which are somehow based on the idea of random search (see, for example, \cite{Ferreiro2013,Garcia2014,Langdon2011}). By virtue of their stochastic nature, algorithms of this type guarantee convergence to the global minimum only in the sense of probability, which differentiates them unfavorably from deterministic methods.

With regard to many deterministic algorithms of Lipschitzian global optimization with guaranteed convergence, parallel variants have been proposed \cite{Evtushenko2009,He2008,Paulavicius2011}. However, these versions of algorithms are parallelized on the CPU using shared and/or distributed memory; presently, no GPU implementations have been made. For example, \cite{Paulavicius2011} describes parallelization of an algorithm based on the ideas of the branch and boundary method using MPI and OpenMP.

Within the framework of this research, we consider the problems of capturing the optimum, which are characterized by a lengthy period for calculating the values of objective function in comparison with the time needed for processing them. For example, objective function can be specified using systems of linear algebraic equations, systems of ordinary differential equations, etc. Currently, graphics accelerators can be used to solve problems of this type. Moreover, an accelerator can solve several such problems at once \cite{Kindratenko2014}; i.e., using the GPU, one can calculate multiple function values simultaneously.

@@

\section{Multidimensional parallel global search algorithm}
Consider the problem of finding the global minimum of the \(N\)-dimensional function \(\varphi(y)\) in the hyperinterval  \(D=\{y\in R^N:a_i\leqslant x_i\leqslant{b_i}, 1\leqslant{i}\leqslant{N}\}\)  We will assume that the function satisfies the Lipschitz condition with an a priori unknown constant \(L\).
\begin{equation}
\label{task}
\varphi(y^*)=\min\{\varphi(y):y\in D\}
\end{equation}
\begin{equation}
\label{lip}
|\varphi(y_1)-\varphi(y_2)|\leqslant L\Vert y_1-y_2\Vert,y_1,y_2\in D,0<L<\infty
\end{equation}

@@

The questions of numerical construction of approximations to Piano curve (evolvents) and the corresponding theory are discussed in detail in \cite{Strongin2000}, \cite{Sergeyev2013}. Using evolvents \(y(x)\) reduces the multidimensional problem (\ref{task}) to a one-dimensional problem
\begin{displaymath}
\label{oneDimTask}
\varphi(y^*)=\varphi(y(x^*))=\min\{\varphi(y(x)):x\in [0,1]\}
\end{displaymath}

An important property is that the relative differences of the function remain limited: if the function \(\varphi(y)\) in the region \(D
\) satisfies the Lipschitz condition, then the function \(\varphi(y(x))\) in the interval \([0,1]\) will satisfy the uniform Holder condition
\begin{displaymath}
\label{holder}
|\varphi(y(x_1))-\varphi(y(x_2))|\leqslant H{|x_1-x_2|}^{\frac{1}{N}},x_1,x_2\in[0,1]
\end{displaymath}
where the Holder constant \(H\) is related to the Lipschitz constant \(L\) by the ratio
\begin{displaymath}
H=4Ld\sqrt{N},d=\max\{b_i-a_i:1\leqslant i\leqslant N\}
\end{displaymath}

Therefore, without limiting generality, we can consider minimizing the one-dimensional function \(f(x)=\varphi(y(x)), x\in[0,1]\), which satisfies the Holder condition.

@@

instead of using a single Peano curve \(y(x)\) (see \cite{Strongin1992}, \cite{Strongin2000}). For example, each Peano curve \(y^i(x)\) from \(Y_L(x)\) can be obtained as a result of some shift along the main diagonal of the hyperinterval \(D\), the original curve \(y(x)\). Another way is to rotate the sweep around the coordinate. The set of evolvents that have been constructed allows us to obtain for any close images \(y', y''\), close prototypes \(x', x''\) for some mapping \(y^i(x)\).

@@

Each problem from this set can be solved independently, and any calculated value \(z= \varphi(y')\), \(y'=y^i(x')\) of the function \(\varphi(y)\) in the \(i\)-th problem can be interpreted as calculating the value \(z= \varphi(y')\), \(y'=y^s(x'')\) for any other \(s\)-th problem without repeated labor-intensive calculations of the function \(\varphi(y)\). Such informational unity makes it possible to solve the entire set of problems in a parallel fashion. This approach was discussed in detail in \cite{Barkalov2019}.

@@

The recursive optimization scheme is based on the well-known relation \cite{Ref22}
\begin{equation}
\label{nestedScheme}
\min{\varphi(y):y\in D}=\min_{a_1\leqslant y_1\leqslant b_1}\min_{a_2\leqslant y_2\leqslant b_2}\dots\min_{a_1\leqslant y_N\leqslant b_N}\varphi(y)
\end{equation}
which allows one to replace the solution of the multidimensional problem (\ref{task}) with the solution of a family of one-dimensional subproblems recursively related to each other.
Let's introduce a set of functions

\begin{equation}
\label{phiN}
\varphi_N(y_1,\dots,y_N)=\varphi(y_1,\dots,y_N)
\end{equation}
\begin{equation}
\label{phiI_}
\varphi_i(y_1,\dots,y_i)=\min_{a_{i+1}\leqslant y_{i+1} \leqslant b_{i+1}}\varphi_{i+1}(y_1,\dots,y_i,y_{i+1}),1\leqslant i\leqslant N-1
\end{equation}

@@

\begin{equation}
\label{phiFirst}
\varphi_1(y_1^*)=\min\{\varphi_1(y_1),y_1\in [a,b]\}
\end{equation}

However, each calculation of the value of the one-dimensional function  \(\varphi_1(y_1)\) at some fixed point corresponds to the solution of a one-dimensional minimization problem

\begin{displaymath}
\label{phi2}
\varphi_2(y_1,y_2^*)=\min\{\varphi(y_1,y_2):y_2\in [a_2,b_2]\}
\end{displaymath}

@@

For the recursive scheme described above, a generalization (block recursive scheme) is proposed that combines the use of evolvents and a recursive scheme in order to efficiently parallelize computations.

Consider the vector $y$ as a vector of block variables
\begin{displaymath}
y=(y_1,\dots,y_N)=(u_1,u_2,\dots,u_M),
\end{displaymath}
where the \(i\)-th block variable \(u_i\) is a vector of sequentially taken components of the vector \(y\), i.e. $u_1=(y_1,y_2,\dots,y_{N_1})$,$u_2=(y_{N_1+1},y_{N_1+2},\dots,y_{N_1+N_2})$,$\dots,u_M=(y_{N-N_M+1},y_{N-N_M+2},\dots,y_N)$, while \(N_1+N_2+\dots+N_M=N\).

@@

\begin{equation}
\label{blockNested}
\min_{y\in D} \varphi(y)=\min_{u_1\in D_1}\min_{u_2\in D_2}\dots\min_{u_M\in D_M}\varphi(y)
\end{equation}

@@

\begin{equation}
\label{subTasks}
\varphi_i(u_1,\dots,u_i)=\min_{u_{i+1}\in D_{i+1}}\varphi_{i+1}(u_1,\dots,u_i,u_{i+1}),1\leqslant i\leqslant M-1
\end{equation}

@@

The obvious method (elaborated in detail in \cite{Grishagin1997,Grishagin2001,Grishagin2015} for the nested optimization scheme and in \cite{Barkalov2014},\cite{Ref27} for the block nested optimization scheme) is based on solving subproblems in accordance with the recursive order of their generation. However, there is a significant loss of information about the target function.

@@

In the case of one-dimensional subproblems, this approach was theoretically substantiated and tested in \cite{Grishagin2016,Grishagin2016_1,Grishagin2018}, and in the article [] a generalization of the adaptive scheme for multidimensional subproblems was proposed.

@@

Let the nested subproblems in the form (\ref{subTasks}) be solved using the global search algorithm described in subsection 1. Then each subproblem (\ref{subTasks}) can be assigned a numerical value called the characteristic of this problem. As such, we can take the maximum characteristic of the intervals formed in this problem – \(R(t)\). In accordance with the rule for calculating characteristics, the higher the value of the characteristic, the more promising the subproblem is in the continued search for the global minimum of the original problem (\ref{task}). Therefore, at each iteration, subproblems with the maximum characteristic are selected for conducting the next test. The test either leads to the calculation of the value of the objective function \(\varphi_1(y)\) (if the selected subproblem belonged to the level \(j=M\)), or generates new subproblems according to (\ref{subTasks}) for \(j \geqslant  M-1\). In the latter case, the newly generated problems are added to the current set of problems, their characteristics are calculated, and the process is repeated. The optimization process is completed when the root problem satisfies the condition for stopping the algorithm that solves this problem. Some results pointing in this direction are presented in \cite{Ref31}, \cite{Ref32}.

@@

\ref{fig:Parallel_Computing} shows the general scheme of organizing calculations using several cluster nodes and several GPUs. In accordance with this scheme, nested subproblems  ${{\varphi }_{i}}({{u}_{1}},...,{{u}_{i}})=\underset{{{u}_{i+1}}\in {{D}_{i+1}}}{\mathop{\min }}\,{{\varphi }_{i+1}}({{u}_{1}},...,{{u}_{i}},{{u}_{i+1}})$  with \(i=1,\ldots ,M-2\) are solved using only the CPU. The values of the optimized function are not calculated directly in these subproblems: the calculation the function value  ${{\varphi }_{i}}({{u}_{1}},...,{{u}_{i}})$ is a solution to the minimization problem at the next level.

The subproblem of the last \((M - 1)\) level \({{\varphi }_{M-1}}({{u}_{1}},...,{{u}_{M-1}})=\underset{{{u}_{M}}\in {{D}_{M}}}{\mathop{\min }}\,{{\varphi }_{M}}({{u}_{1}},...,{{u}_{M}})\) differs from all the previous subproblems – it calculates the values of the optimized function, since ${{\varphi }_{M}}({{u}_{1}},...,{{u}_{M}})=\varphi ({{y}_{1}},...,{{y}_{N}})$. This subproblem transfers between the CPU and the GPU.

@@

In the study \cite{Gaviano2003}, \cite{Ref16}, a GKLS generator is described that allows one to generate multiextremal optimization problems with previously known properties: the number of local minima, the size of their regions of attraction, the global minimum point, the value of the function in it, etc.

Below are the results of a numerical comparison of three sequential algorithms -- DIRECT \cite{Jones2009}, DIRECT\textit{l} \cite{Gablonsky2001} and Global Search Algorithm (GSA); (results of the first two algorithms are given in the article \cite{Sergeyev2006}). A numerical comparison was carried out on the Simple and Hard function classes of dimension 4 and 5 from \cite{Sergeyev2006}. The Global minimum $y*$ was considered found if the algorithm generated a test point $y^k$ in the $\delta$-neighborhood of the global minimum, i.e. $\left\| {{y}^{k}}-{{y}^{*}} \right\|\le \delta $. Also, the size of the neighborhood was chosen (in accordance with \cite{Sergeyev2006}) as $\delta =\left\| b-a \right\|\sqrt[N]{\Delta }$, here $N$ is the dimension of the problem to be solved, a и b are the boundaries of the search domain $D$, the parameter  $\Delta $=10-6 for N=4 and $\Delta =10-7$ for $N=5$. When using the AGP method for the Simple class, the parameter $r=4.5$ was selected, for the Hard class - $r=5.6$; the parameter for constructing the Peano curve was fixed, $m=10$. The maximum number of iterations allowed was $K_{max} = 10^6$.

\begin{table}
    \centering
    \begin{tabular}{|c|p{3cm}|c|c|c|}
    \hline
    \(N\) & Problem class & DIRECT & DIRECT\(l\) & GSA \\ \hline
    4 & Simple & >47282 (4) & 18983 & 11953\\ \cline{2-5}
      & Hard & >95708 (7) & 68754 & 25263\\ \hline
    5 & Simple &>16057 (1) & 16758 & 15920\\ \cline{2-5}
      & Hard & >217215 (16) & >269064 (4) & >148342 (4)\\ \hline
    \end{tabular}
    \caption{Average number of iterations}
    \label{table:average_iters}
\end{table}

   Table \ref{table:average_iters} shows the average number of iterations, $k_{av}$, that the method performed when solving a series of problems from these classes. The symbol ">" reflects a situation where not all problems of the class were solved by any method whatsoever. This means that the algorithm was halted because the maximum allowed number of $K_{max}$ iterations was reached. In this case, the value $K_{max} = 10^6$ was used to calculate the average value of the number of iterations, $k_{av}$, which corresponds to the lower estimate of this average value. The number of unsolved problems is indicated in parentheses.

  As can be seen from Table \ref{table:average_iters}, the sequential GSA surpasses the DIRECT and DIRECT\textit{l} methods in all classes of problems in terms of the average number of iterations. At the same time, in the 5-Hard class, none of the methods solved all the problems: DIRECT failed to solve 16 problems, DIRECTl and GSA -- 4 problems each.

@@

%Следующий эксперимент будет проведен на 6-ти мерных задачах GKLS. В таблице \ref{table:GKLS6_average_MMAS_GPU} приведено ускорение по времени и по итерациям полученное при решении серии задач на GPU с использованием алгоритма объединяющего множественные развертки и адаптивную схему по сравнению с адаптивной схемой на CPU использующей 4 потока. Эксперимент проводился на трех узлах кластера и спользовалось до 6144 нитей.

@@

%
% ---- Bibliography ----
%
\bibliographystyle{spmpsci}
\bibliography{bibliography}{}

%\begin{thebibliography}{6}
%
%\bibitem {Ref1}	Hwu, W. GPU Computing Gems: Emerald Edition — San Francisco: Morgan Kaufmann, 2011. — 886 p.
%\bibitem {Ref2}	A.M. Ferreiro, J.A. Garcia, J.G. Lopez-Salas, C. Vazquez. An efficient implementation of parallel simulated annealing algorithm in GPUs // Journal of global optimization. — 2013. — V. 57, No 3. — P. 863–890.
%\bibitem {Ref3}	J.M. Garcia-Martinez, E.M. Garzon, P.M. Ortigosa. A GPU implementation of a hybrid evolutionary algorithm: GPuEGO // Journal of super-computing. — 2014. — V. 70, No 2. — P. 684–695.
%\bibitem {Ref4}	Langdon, W.B. Graphics processing units and genetic programming: an overview // Soft Computing. — 2011. — V. 15, No 8. — P. 1657–1669.
%\bibitem {Ref5}	Ю.Г. Евтушенко, В.У. Малкова, А.А. Станевичюс. Параллельный поиск глобального экстремума функций многих переменных // Ж. вычисл. матем. и матем. физ. — 2009. — Т. 49, No2. — С. 255–269.
%\bibitem {Ref6}	J. He, A. Verstak, L.T. Watson, M. Sosonkina. Design and implementation of a massively parallel version of DIRECT // Computational optimization and applications. — 2008. — V. 40, No 2. — P. 217–245.
%\bibitem {Ref7}	R. Paulavicius, J. Žilinskas, A. Grothey. Parallel branch and bound for global optimization with combination of Lipschitz bounds // Optimization methods and software. — 2011. — V. 26, No 3. — P. 487–498.
%\bibitem {Ref8}	V. Kindratenko (ed.). Numerical computations with GPUs — New York: Springer, 2014. — 405 p.
%\bibitem {Ref9}	Strongin, R.G. Global Optimization with Non-convex Constraints. Sequential and Parallel Algorithms / R.G. Strongin, Ya.D. Sergeyev – Kluwer Academic Publishers, 2000. – 704 p.
%\bibitem {Ref10}	Sergeyev, Ya.D. Introduction to global optimization exploiting space-filling curves / Ya.D. Sergeyev, R.G. Strongin, D. Lera – Springer, 2013. – 125 p.
%\bibitem {Ref11}	Стронгин, Р.Г. Параллельные вычисления в задачах глобальной оптимизации / Р.Г. Стронгин, В.П. Гергель, В.А. Гришагин, К.А. Баркалов – М.: Издательство Московского университета, 2013. – 280 с.
%\bibitem {Ref12}	Barkalov, K.A. A global optimization technique with an adaptive order of checking for constraints / K.A. Barkalov, R.G. Strongin // Computational Mathematics and Mathematical Physics. – 2002. – Vol. 42, No. 9. – P. 1289-1300.
%\bibitem {Ref13}	Gergel, V.P. A method of using derivatives in the minimization of multiextremum functions / V.P. Gergel // Computational Mathematics and Mathematical Physics. – 1996. – Vol. 36, No. 6. – P. 729 – 742.
%\bibitem {Ref14}	Gergel, V.P. A global optimization algorithm for multivariate functions with lipschitzian first derivatives / V.P. Gergel // Journal of Global Optimization. – 1997. – Vol. 10, No. 3. – P. 257-281.
